Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer

### 2024-10-24 manuscript.qmd

$$
P \big( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \mid d, \boldsymbol{\hat{\theta}} \big)
= \frac{P \left( \boldsymbol{\zeta}^\text{i}_k \mid \mathbf{X}^\text{i} \right) P \left( \boldsymbol{\zeta}^\text{c}_\ell \mid \mathbf{X}^\text{c} \right) P \big( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \mid \boldsymbol{\hat{\theta}}, \text{T}x \big)}
{\sum_{i=0}^{2^V} \sum_{j=0}^{2^V} P \left( \boldsymbol{\zeta}^\text{i}_k \mid \mathbf{X}^\text{i}=\boldsymbol{\xi}^\text{i}_i \right) P \big( \boldsymbol{\zeta}^\text{c}_\ell \mid \mathbf{X}^\text{c}=\boldsymbol{\xi}^\text{c}_j \big) P \big( \mathbf{X}^\text{i}=\boldsymbol{\xi}^\text{i}_i, \mathbf{X}^\text{c}=\boldsymbol{\xi}^\text{c}_j, \epsilon \mid \boldsymbol{\hat{\theta}}, \text{T}x \big)}
$$ {#eq-uni-bayes-law}

@@

![Visual representation of @eq-bilateral-marginal-algebra for the case of midline extention. The left and right matrices represent the evolution of the possible hidden states for the ipsi- and contralateral neck respectively. In particular, the right matrices describe the evolution of the contralateral hidden states once for the case of no midline extension (top) and once for the case where the tumor does extend over the mid-sagittal line (bottom). In the center, the time-prior for late T-category is plotted as a diagonal matrix. The result of this matrix computation is the joint distribution $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)$](figures/fig_model_algebra_equation.svg){#fig-model-algebra-equation}

@@

![The full distribution $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\text{False} \right)$ over the ipsi- and contralateral states, as well as midline extension as computed by a model considering the LNLs II, III, and IV. Since this is a three-dimensional object with one dimension being the binary random variable representing midline extension $\epsilon$, we show this distribution as separate two-dimensional heatmaps. These two matrices are the result of @eq-bilateral-marginal-algebra, which we visualized as well in @fig-model-algebra-equation.](figures/fig_model_state_dist.svg){#fig-model-state-dist}

>>>>

$$
P \big( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \mid d, \boldsymbol{\hat{\theta}} \big)
= \frac{P \left( \boldsymbol{\zeta}^\text{i}_k \mid \mathbf{X}^\text{i} \right) P \left( \boldsymbol{\zeta}^\text{c}_\ell \mid \mathbf{X}^\text{c} \right) P \big( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \mid \boldsymbol{\hat{\theta}}, \text{T}x \big)}
{\sum_{i=0}^{2^V} \sum_{j=0}^{2^V} \mathcal{C}_{ij}}
$$ {#eq-uni-bayes-law}

with the normalization constants

$$
\mathcal{C}_{ij} = P \left( \boldsymbol{\zeta}^\text{i}_k \mid \mathbf{X}^\text{i}=\boldsymbol{\xi}^\text{i}_i \right) P \big( \boldsymbol{\zeta}^\text{c}_\ell \mid \mathbf{X}^\text{c}=\boldsymbol{\xi}^\text{c}_j \big) P \big( \mathbf{X}^\text{i}=\boldsymbol{\xi}^\text{i}_i, \mathbf{X}^\text{c}=\boldsymbol{\xi}^\text{c}_j, \epsilon \mid \boldsymbol{\hat{\theta}}, \text{T}x \big)
$$

@@

![Visual representation of @eq-bilateral-marginal-algebra for the case of midline extention. The left and right matrices represent the evolution of the possible hidden states for the ipsi- and contralateral neck respectively. In particular, the right matrices describe the evolution of the contralateral hidden states once for the case of no midline extension (top) and once for the case where the tumor does extend over the mid-sagittal line (bottom). In the center, the time-prior for late T-category is plotted as a diagonal matrix. The result of this matrix computation is the joint distribution $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)$, visualized in @fig-model-state-dist](figures/fig_model_algebra_equation.svg){#fig-model-algebra-equation}

@@

![The full distribution $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\text{False} \right)$ over the ipsi- and contralateral states, as well as midline extension as computed by a model considering the LNLs II, III, and IV for a late T-category tumor. Since this is a three-dimensional object with one dimension being the binary random variable representing midline extension $\epsilon$, we show this distribution as separate two-dimensional heatmaps. These two matrices are the result of @eq-bilateral-marginal-algebra, which we visualized as well in @fig-model-algebra-equation.](figures/fig_model_state_dist.svg){#fig-model-state-dist}
